A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2022-05-04 submission/manuscript.rmd
updating manuscript with new results

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```
\thispagestyle{empty}
\vspace{8mm}

@@

Understanding how a treatment’s effect varies across patients---a concept
described as heterogeneity of treatment effects (HTE)---has been central to the
agenda for both personalized (or precision) medicine and comparative
effectiveness research. More formally, HTE has been defined as the non-random
variability in the direction or magnitude of a treatment effect, in which the
effect is measured using clinical outcomes [@KRAVITZ2004]. Usually, analyses
focus on the relative scale, where treatment effects are assessed one at a time
in patient subgroups defined from single covariates, an approach that suffers
from low power and multiplicity issues [@Yusuf1991; @FernandezyGarcia2010].
However, even with well-established constant relative effects, treatment benefit
(or harm) may vary substantially on the absolute scale.

In recent years, a large number of methods has been developed for the
assessment of HTE, mainly in the RCT setting. Earlier work suggested separating
HTE analyses into exploratory, confirmatory, descriptive and predictive
[@Varadhan2013]. Exploratory analyses focus on hypothesis generation,
confirmatory analyses test subgroup effect hypotheses, descriptive analyses aim
at facilitating future synthesis of subgroup effects and predictive analyses
predict probabilities of benefit or harm in individual patients. Predictive HTE
approaches can be further subdivided into risk modeling, treatment effect
modeling and optimal treatment regime methods, based on the reference class used
for defining patient similarity when making individualized predictions or
recommendations [@Rekkas2020]. We focus on “risk modeling” approaches where
patients are divided into risk strata using either an existing or an internally
developed risk prediction model. Risk-stratum-specific estimates provide an
overview of the evolution of treatment effects with increasing risk both on the
relative and the absolute scale. Recently, systematic guidance on the
application of such methods has been developed [@Kent2010; @Kent2018].

While these approaches were developed for application in randomized controlled
trials (RCTs), observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, including heterogeneous populations. However, unlike RCTs,
treatment effects are subject to confounding, while the unique structure of
different databases calls for database-specific analysis plans that often are
not easily transportable. Because of the latter issue, running analyses at scale
demands a big investment of time and effort, as researchers are forced to map
their analysis plans to the databases available to them.

The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
established an international network of data partners and researchers that aim
to bring out the value of health data through large-scale analytics by mapping
all available databases to the Observational Medical Outcomes Partnership (OMOP)
Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012].

We aimed to develop a framework for implementing a risk-based predictive
approach for evaluating HTE in high-dimensional observational data, extending
the existing guidelines of the RCT setting. Our publicly available package
provides an out-of-the-box solution for implementing such analyses at scale
within the OHDSI network, taking advantage of the OMOP-CDM. We implemented the
framework using existing OHDSI methods including the patient-level prediction
framework and the population-level effect estimation framework based on new-user
cohort design [@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed
heterogeneity of the effects of first-line hypertension treatment: we compared
the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers
on 9 outcomes across three different US claims databases.

# Materials and Methods

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation of the results. We developed an open-source R-package for the
implementation of the proposed framework and made it publicly available. The
source code can be found at <https://github.com/OHDSI/RiskStratifiedEstimation>.

## Step 1: General definition of the problem

The typical research aim is: "to compare the effect of treatment $T$ to a
comparator treatment $C$ in patients with disease $D$ with respect to outcomes
$O_1,\dots,O_n$".

Cohort definitions are crucial for this step of the framework. We define a
cohort as the set of patients who satisfy one or more inclusion criteria for a
duration of time. A cohort within the OHDSI setting is more than a set of
specific clinical codes, providing a definition of a logic for how to use that
code set. All cohort definitions consist of: an entry event, i.e. the time a
patient enters a cohort; a set of inclusion criteria applied to the initial
event cohort to further restrict the set of people, resulting in the
construction of the construction of the qualifying cohort; cohort exit criteria
that terminate the patient's presence in the cohort. Cohort definitions are
transportable, meaning that in theory they can be implemented in any database,
provided that it is mapped to the OMOP-CDM.

Our framework uses a comparative cohort design. This means that at least 3
cohorts of patients need to be defined at this stage of the framework.

  - A single treatment cohort ($T$) which includes patients with disease $D$
    receiving the target treatment of interest. For example, a set of
    hypertension patients within a database that receive angiotensin-converting
    enzyme inhibitors, followed from the time of initiation until the time of
    censoring.
  - A single comparator cohort ($C$) which includes patients with disease $D$
    receiving the comparator (control) treatment. For example, a set of patients
    in a database that receive beta blockers, followed from the time of
    initiation until the time of censoring.
  - One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
    developing the outcomes of interest. For example, the set of patients in a
    database that have at least one occurrence of acute myocardial infarction
    (MI) in their record.

## Step 2: Identification of the database

The aim of this step is the inclusion of databases that represent the patient
population of interest. It is required that the databases are mapped to the
OMOP-CDM. The inclusion of multiple databases potentially increases the
generalizability of results. Furthermore, the cohorts should preferably have
adequate sample size with adequate follow-up time to ensure precise effect
estimation, even within smaller risk strata. Other issues that may be of
importance for database inclusion are the depth of data capture (the precision
at which measurements, lab tests, conditions are recorded), the reliability of
data entry and many more, also depending on the task at hand.

@@

We adopt the standardized framework for the generation of patient-level
prediction models using observational data that ensures adherence to existing
guidelines [@Collins2015; @Moons2015]. This prediction framework requires the
definition of two essential cohorts: a target cohort, i.e. a set of patients
that satisfy one or more inclusion criteria for a duration of time, and an
outcome cohort.

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. Further restrictions can be applied on the target cohort to
construct the final population on which the prediction model will be developed
(e.g. exclude patients with a prior outcome in their history, before being
included in the target cohort). To avoid deferentially fitting the prediction
model to patients across treatment arms, thus introducing spurious interactions
with treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
prediction model in the propensity score-matched (1:1) subset of the population.

More specifically, we first estimate propensity scores using LASSO logistic
regression and a large set of baseline covariates including demographics, drug
exposures, diagnoses, measurements and medical devices. We match patients 1-1
using a caliper, i.e. the maximum distance that is acceptable for any match. The
default value we use is 0.2 on the standardized logit scale for the propensity
scores. Other methods of fitting the propensity scores, such as random forest
and others can also be considered.

Finally, we need to define the time horizon within which we aim to make
predictions and we also need to select the machine-learning algorithm we want to
use to generate patient-level predictions. Currently, the available options are
regularized logistic regression, random forest, gradient boosting machines,
decision tree, naive Bayes, K-nearest neighbors, neural network and deep
learning (convolutional neural networks, recurrent neural network and deep
nets).

After model development, a performance overview of the derived prediction models
including discrimination and calibration both in the propensity score matched
subset, the entire population and separately for treated and comparator patients
should also be reported. This is important to ensure that no overfitting of the
prediction model in one of the cohorts has occurred. In addition, the
performance of the prediction models is directly related to our ability to
single out patient subgroups where treatment may be highly beneficial or unsafe.
Kent et al [@Kent2016] demonstrated that the event rate and the discriminative
ability of the prediction model can predict very well the distribution of
predicted risk. Lower event rate and higher c-statistic (given good calibration)
result in high risk heterogeneity, thus making estimated average treatment
effects uninformative. In this case, risk stratified analysis of HTE can be more
effective in singling out patient subgroups that stand to benefit (or be harmed)
most by treatment in question.

The aim of this step is the estimation of treatment effects (both on the
relative and the absolute scale) within risk strata---typically 4 risk
quarters---defined using the prediction model of step 3. Effect estimation may
be focused on the difference in outcomes for a randomly selected person from the
risk stratum (average treatment effect) or for a randomly selected person from
the treatment cohort within the risk stratum receiving the treatment under study
(average treatment effect on the treated).

Any appropriate method for the evaluation of relative and absolute treatment
effects can be considered, as long as the this is done consistently in all risk
strata. Common approaches are odds ratios or hazard ratios for relative scale
estimates and differences in observed proportions or differences in Kaplan-Meier
estimates for absolute scale estimates, depending on the problem at hand. We
estimate propensity scores within risk strata which we then use to match
patients from different treatment cohorts or stratify them into groups with
similar propensity scores or to weigh each patient's contribution to the
estimation process [@Austin2011].

Before focusing on the results of the estimation process we need to evaluate if
adequate covariate balance was achieved within each risk stratum accounting for
measured confounding. Common approaches include evaluation of the overlap of
propensity score distributions and calculation of standardized covariate
differences before and after propensity score adjustment.

A schematic overview of the prediction and estimation steps is presented in
Figure \@ref(fig:figure1).

```{r figure1, echo=FALSE, fig.cap="(A) Starting from a treatment (top), a comparator (bottom) and an outcome (middle) cohort we estimate the propensity scores on the entire target population. (B) We match patients on the propensity scores and estimate the prediction model. Since we match patients we develop the prediction model on smaller subset of the initial population and, therefore, the number of patients is smaller in B compared to A. (C) We apply the prediction model on the entire population (green: lower 25\\% of the risk distribution; yellow: patients with risk between 25\\% and 50\\% of the risk distribution; orange: patients with risk between 50\\% and 75\\% of the risk distribution; red: patients at risk higher than 75\\% of the risk distribution). (D) We separate in risk subgroups, here quarters. Within risk quarters propensity scores are estimated again and relative and absolute treatment effects are estimated.", out.width = '100%'}
# knitr::include_graphics(here("figures/Figure_1.tif"))
grid::grid.raster(tiff::readTIFF(here("figures/Figure_1.tif")))
```

As a proof of concept, we focus on the comparison of angiotensin converting
enzyme (ACE) inhibitors to beta blockers are among the most common treatment
classes for hypertension, with well-established effectiveness. Beta blockers,
even though initially widely used for the treatment of hypertension, more recent
trials and meta-analyses have cast doubt on their relative effectiveness
[@wiysonge2017beta]. As a result, newer US guidelines do not consider beta
blockers for initial treatment for hypertension while in the EU guidelines
combination with other antihypertensive treatments is recommended
[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
the efficacy profile of beta blockers is similar to other major treatment
classes in younger hypertensive patients and, thus, countries like Canada still
include them as a first-line treatment candidate [@Khan2006; @Rabi2020].

## Step 1: General definition of the problem

We consider the following research aim: "compare the effect of ACE-inhibitors
($T$) to the effect of beta blockers ($C$) in patients with established
hypertension ($D$) with respect to 9 outcomes ($O_1,\dots,O_9$)". The cohorts
are:

  - Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
    with at least one year of follow-up before treatment initiation and a
    recorded hypertension diagnosis within that year.
  - Comparator cohort: Patients receiving any drug within the beta blocker class
    with at least one year of follow-up before treatment initiation and a
    recorded hypertension diagnosis within that year.
  - Outcome cohorts: We consider 3 main and 6 safety outcome cohorts. These are
    patients in the database with a diagnosis of: acute MI; hospitalization with
    heart failure; ischemic or hemorrhagic stroke (efficacy outcomes);
    hypokalemia; hyperkalemia; hypotension; angioedema; cough; abnormal weight
    gain (safety outcomes).

All cohort definitions were identical to the ones used in the multinational
study carried out within OHDSI that provided overall treatment effect estimates
comparing all anti-hypertensive drug classes with each other [@Suchard2019].
More information can be found in the supplementary material.

We used the following databases:

  - IBM MarketScan Medicare Supplemental Beneficiaries (MDCR): Represents health
    services of retirees (aged 65 or older) in the United States with primary or
    Medicare supplemental coverage through privately insured fee-for-service,
    point-of-service or capitated health plans. These data include adjudicated
    health insurance claims (e.g. inpatient, outpatient and outpatient
    pharmacy). Additionally, it captures laboratory tests for a subset of the
    covered lives.
  - IBM MarketScan Medicaid (MDCD): Adjudicated US health insurance claims for
    Medicaid enrollees from multiple states. It includes hospital discharge
    diagnoses, outpatient diagnoses and procedures and outpatient pharmacy
    claims as well as ethnicity and Medicare eligibility.
  - IBM MarketScan Commercial Claims and Encounters (CCAE): Data from
    individuals enrolled in US employer-sponsored insurance health plans. The
    data includes adjudicated health insurance claims (e.g. inpatient,
    outpatient, and outpatient pharmacy) as well as enrollment data from large
    employers and health plans who provide private healthcare coverage to
    employees, their spouses and dependents. Additionally, it captures
    laboratory tests for a subset of the covered lives.

Our analyses included a total of 784,561, 66,820 and 101,661 patients initiating
treatment with ACE inhibitors and 395,740, 45,999 and 69,798 patients initiating
treatment with beta blockers in CCAE, MDCD and MDCD respectively (Table
\@ref(tab:table)). Adequate numbers of patients were included in all strata of
predicted acute MI risk.

\captionsetup{width=\textwidth}
```{r table, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, results='asis', booktabs=TRUE}
table <- readr::read_csv(
  here::here("data/processed/table1.csv"),
  na = character()
) %>%
  select(-database)

knitr::kable(
  table,
  format = "latex",
  longtable = TRUE,
  booktabs = TRUE,
  col.names = c(
    "Outcome", "Risk quarter",
    "Persons", "Person years", "Events",
    "Persons", "Person years", "Events"
  ),
   format.args = list(
    big.mar = ","
  ),
  caption = "Number of patients, person years and events within quarters of predicted risk for acute MI for the 3 main outcomes of the study (acute MI, hospitalization with heart failure and ischemic or hemorrhagic stroke) across the 3 databases. The populations used for the evaluation of hosptialization with heart failure and stroke are subsets of the populations used for the evaluation of actute MI."
) %>%
  kableExtra::group_rows("CCAE", 1, 12) %>%
  kableExtra::group_rows("MDCD", 13, 24) %>%
  kableExtra::group_rows("MDCR", 25, 36) %>%
  kableExtra::add_header_above(
    c(
      "", "", 
      "ACE inhibitors" = 3,
      "Beta blockers"  = 3
    )
  )
```

We internally developed separate prediction models for acute MI in all 3
databases.

The prediction models were estimated on the propensity score matched (1:1)
subset of the population, using caliper of 0.2 and after excluding patients
having the outcomes any time prior to treatment initiation. We chose a time
horizon of 2 years after inclusion into the target cohort. For this
demonstration, we developed the prediction models using LASSO logistic
regression with 3-fold cross validation for hyper-parameter selection. For this
demonstration, we developed the prediction models using LASSO logistic
regression with 3-fold cross validation for hyper-parameter selection.

The models developed in the 3 databases had moderate discriminative performance
(internally validated) with no major issues of overfitting to any cohort except
for the case of CCAE database in which the derived prediction model performed
better in the comparator cohort (Table \@ref(tab:tableAucs)). We also observed
lower performance of the prediction model developed in MDCR compared to the
other 2 databases. Results on the calibration of the prediction models can be
found in the supplement.

\captionsetup{width=\textwidth}
```{r tableAucs, cache=TRUE, echo=FALSE, results='asis', booktabs=TRUE}
table <- tibble(
  population = c(
    "Matched set",
    "Treatment cohort",
    "Comparator cohort",
    "Entire population"
  ),
  ccae = c(
    75.78, 70.65, 79.1, 74.06
  ),
  mdcd = c(
    78.12, 78.29, 80.26, 79.46
  ),
  mdcr = c(
    71.09, 69.03, 68.53, 69.07
  )
)

knitr::kable(
  table,
  format = "latex",
  longtable = TRUE,
  booktabs = TRUE,
  col.names = c(
    "Population", "CCAE", "MDCD", "MDCR"
  ),
   format.args = list(
    big.mar = ","
  ),
  caption = "Discriminiative ability (c-statistic) of the derived prediction models for acute myocardial infarction in the matched set (development set), the treatment cohort, the comparator cohort and the etnire population in CCAE, MDCD and MDCR"
)
```

@@

We used patient-level predictions to stratify the patient population into 4 risk
quarters. Within risk strata, relative effects were estimated using Cox
regression and absolute effects were estimated from the Kaplan-Meier estimate
differences at 2 years after treatment initiation. To adjust for observed
confounding within risk strata, we estimated propensity scores using the same
approach as in the development of prediction models. We used the estimated
propensity scores to stratify patients into 5 strata, within each risk quarter.

In general, there was sufficient overlap of propensity score distribution in all
risk strata (Figure \@ref(fig:psDensity)). If empirical equipoise was not
achieved, the validity of the comparative effectiveness estimates would be
questionable.

```{r psDensity, cache = TRUE, echo=FALSE, fig.cap="Preference score distributions for the evaluation of heterogeneity of the effect of ACE inhibitors compared to beta blockers on acute MI based on quarters of predicted acute MI risk. The preference score is a transformation of the propensity score that adjusts for prevalence differences between populations.", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/PsDensity.tiff")))
```

Propensity score adjustment achieved balance for most of the considered
covariates, measured using standardized mean differences before and after
adjustment (Figure \@ref(fig:covariateBalance)). However, in lower risk strata
imbalances persisted for a substantial subset of the covariates, all related to
pregnancy findings. This was anticipated (use of ACE inhibitors is specifically
contraindicated during pregnancy) and was already pointed out in
[@Suchard2019].

```{r covariateBalance, cache = TRUE, echo=FALSE, fig.cap="Patient characteristic balance for ACE inhibitors and beta blockers before and after stratification on the propensity scores. Each dot represents the standardized difference of means for a single covariate before (x-axis) and after (y-axis) stratification", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/CovariateBalance.tiff")))
```

@@

For hospitalization with acute MI there was an increasing trend in favor ACE
inhibitors compared to beta blockers on the relative scale (hazard ratios
decreased) with increasing acute MI risk. More specifically, hazard ratios
decreased from 0.98 (0.77 to 1.26; 95% CI), 1.30 (0.51 to 3.22; 95% CI) and 1.03
(0.82 to 1.29; 95% CI) to 0.76 (0.71 to 0.82; 95% CI), 0.94 (0.82 to 1.07; 95%
CI) and 1.03 (0.93 to 1.15; 95% CI) in CCAE, MDCD and MDCR respectively (Figure
\@ref(fig:combinedRelative)). In terms of hospitalization with heart failure
relative treatment effect estimates favored ACE inhibitors across all risk
strata in all databases. We found no differences between the two treatments in
their effect on stroke on the relative scale. In terms of the safety outcomes we
found an increased risk of cough and angioedema on the relative scale across all
risk strata. In the case of cough, this effect decreased with increasing risk of
acute MI---from 1.37 (1.33 to 1.41; 95% CI), 1.35 (1.24 to 1.48; 95% CI)
and 1.37 (1.29 to 1.45; 95% CI) in the lowest acute MI risk quarter to 1.26
(1.22 to 1.29; 95% CI), 1.07 (1.00 to 1.14) and 1.10 (1.04 to 1.17; 95% CI) in
the highest acute MI risk quarter in CCAE, MDCD and MDCR respectively.

```{r combinedRelative, cache=TRUE, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment on the relative scale (hazard ratios) within strata of predicted risk of acute MI. Values below 1 favor ACE inhibitors, while values above 1 favor beta blockers.", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/CombinedRelative.tiff")))
```

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
inline_hook <- function(x) {
  if (is.numeric(x)) {
    if (abs(x - round(x)) < .Machine$double.eps) {
      formatted <- format(x, digits = 1, big.mark = ",")
    } else {
      formatted <- format(x, digits = 1 ,nsmall = 2, big.mark = ",")
    }
  } else {
    formatted <- x
  }

  return(formatted)
}

knit_hooks$set(inline = inline_hook)
```
\thispagestyle{empty}
\vspace{8mm}

@@

The Observational Health Data Science and Informatics (OHDSI) collaborative has
established a global network of data partners and researchers that aim to bring
out the value of health data through large-scale analytics by mapping local
databases to the Observational Medical Outcomes Partnership (OMOP) Common Data
Model (CDM) [@hripcsak2015observational;@Overhage2012]. A standardized framework
applying current best practices for comparative effectiveness studies within the
OHDSI setting has been proposed [@Ryan2013]. This framework was implemented on a
large scale for estimation of average effects of all first-line hypertension
treatment classes on a total of 52 outcomes of interest across a global network
of 9 observational databases [@Suchard2019].

Treatment effects can often vary substantially across individual patients,
causing overall effect estimates to be inaccurate for a significant proportion
of the patients at hand [@Rothwell1995;@KRAVITZ2004]. Understanding
heterogeneity of treatment effects (HTE) has been central to the agenda for both
personalized (or precision) medicine and comparative effectiveness research,
giving rise to a wide range of approaches for its discovery, evaluation and
application in clinical practice. While exploratory and confirmatory HTE
analyses focus on generating and testing hypotheses on subgroup effects,
predictive HTE analyses use all available patient data to predict the benefits
or harms of treatment in individual patients [@Varadhan2013;@Kent2018].

Baseline risk has been recognized as a robust predictor of treatment effect,
because of the inherent relationship between baseline risk and the absolute
effects of treatment
[@Rothwell2005;@Hayward2006;@Kent2007;@Kent2008;@Kent2010]. For example, an
invasive coronary procedure – in comparison with medical treatment – improves
survival in patients with myocardial infarction at high (predicted) baseline
risk but not in those at low baseline risk [@Thune2005]. It has also been shown
that high-risk patients with pre-diabetes benefit substantially more from a
lifestyle modification program than low-risk patients [@Sussman2015].

Recently, systematic guidance on the application of risk-based methods has been
developed for RCT data [@Kent2019;@PathEnE]. Patients are divided into risk strata using
either an existing or an internally developed risk prediction
model. Risk-stratum-specific estimates provide an overview of the evolution of
treatment effects with increasing risk both on the relative and the absolute
scale. Several methods for predictive HTE analysis have been adapted for use in
observational data, but risk-based methods for predictive HTE in observational
data are still lacking and have been highlighted as an important future research
need [@PathEnE].

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in observational healthcare databases, extending the existing
methodology developed for the RCT setting. We implemented the framework in a
publicly available package providing an out-of-the-box solution for implementing
such analyses at scale within any observational database mapped to OMOP-CDM. In
a case study we analyzed heterogeneity of the effects of first-line hypertension
treatment: we compared the effect of angiotensin converting enzyme (ACE)
inhibitors to beta blockers on 9 outcomes across three different US claims
databases.

# Methods

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the databases within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation of the results. We developed an open-source R-package for the
implementation of the proposed framework and made it publicly available. The
source code can be found at https://github.com/OHDSI/RiskStratifiedEstimation.

## Step 1: General definition of the research aim

The typical research aim is: “to compare the effect of treatment to a comparator
treatment in patients with disease with respect to outcomes $O_1,\dots,O_n$”.

Our framework uses a comparative cohort design. This means that at least 3
cohorts of patients need to be defined at this stage of the framework:

- A single treatment cohort ($T$) which includes patients with disease receiving
  the target treatment of interest.
- A single comparator cohort ($C$) which includes patients with disease receiving
  the comparator treatment.
- One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
  developing the outcomes of interest

## Step 2: Identification of the databases {#methods2}

Including in our analyses multiple databases representing the population of
interest potentially increases the generalizability of results. Furthermore, the
cohorts should preferably have adequate sample size with adequate follow-up time
to ensure precise effect estimation, even within smaller risk strata. Other
issues that may be of importance for database inclusion are the depth of data
capture (the precision at which measurements, lab tests, conditions are
recorded) and the reliability of data entry.

## Step 3: Prediction {#methods3}

We adopt the standardized framework for the generation of patient-level
prediction models using observational data that ensures adherence to existing
guidelines [@Collins2015;@Moons2015]. To generate the target cohort (i.e. the
set of patients on which the prediction model will be developed), we pool the
already defined treatment cohort and comparator cohort. To avoid deferentially
fitting the prediction model to patients across treatment arms, thus introducing
spurious interactions with treatment [@Burke2014;@vanKlaveren2019], we develop
the patient-level prediction model in the propensity score-matched (1:1) subset
of the sample. Finally, we need to define the time horizon within which we aim
to make predictions and we also need to select the machine-learning algorithm we
want to use to generate patient-level predictions.

After model development, a performance overview of the derived prediction models
including discrimination and calibration both in the propensity score matched
subset, the entire sample and separately for treated and comparator patients
should also be reported.

## Step 4: Estimation {#methods4}

We estimate treatment effects (both on the relative and the absolute scale)
within risk strata—typically 4 risk quarters—defined using the prediction model
of step 3. Effect estimation may be focused on the difference in outcomes for a
randomly selected person from the risk stratum (average treatment effect) or for
a randomly selected person from the treatment cohort within the risk stratum
receiving the treatment under study (average treatment effect on the treated).

Any appropriate method for the analysis of relative and absolute treatment
effects can be considered, as long as the this is done consistently in all risk
strata. Common statistical metrics are odds ratios or hazard ratios for relative
scale estimates and differences in observed proportions or differences in
Kaplan-Meier estimates for absolute scale estimates, depending on the problem at
hand. We estimate propensity scores within risk strata which we then use to
match patients from different treatment cohorts or to stratify them into groups
with similar propensity scores or to weigh each patient’s contribution to the
estimation process [@Austin2011].

Prior to analyzing the results, we need to assess if adequate covariate balance
was achieved within each risk stratum accounting for measured
confounding. Common approaches include analysis of the overlap of propensity
score distributions and calculation of standardized mean differences of the
covariates before and after propensity score adjustment.

A schematic overview of the prediction and estimation steps is presented in
Figure XX.

## Step 5: Presentation of results {#methods5}

## Case study

As a demonstration of our framework, we evaluated if our proposed method was
able to identify treatment effect heterogeneity of ACE inhibitors compared to
beta blockers using acute myocardial infarction (MI) risk quarter specific
effect estimates, both on the relative and on the absolute scale. We focused on
3 efficacy outcomes (acute MI, hospitalization with heart failure and ischemic
or hemorrhagic stroke) and 6 safety outcomes (hypokalemia, hyperkalemia,
hypotension, angioedema, cough and abnormal weight gain). We used data from 3
US-based claims databases. The analysis plan was the framework outlined in
steps 1 through 5.

A recent meta-analysis has shown that beta-blockers are on average less
efficacious in preventing cardiovascular events than ACE inhibitors
[@Wiysonge2017]. However, meta-analyses focusing on age of the patients have
shown that beta blockers have similar efficacy in younger hypertensive patients,
but should not be used as first-line treatment of hypertension in older patients
mainly due to their increased risk of cardiovascular events compared to other
first-line antihypertensive treatments [@Khan2006;@Vgele2017].Our aim was to
replicate these conclusions in our analyses.

## Step 1: General definition of the research aim

We considered the following research aim: “compare the effect of ACE-inhibitors
($T$) to the effect of beta blockers ($C$) in patients with established hypertension
with respect to 9 outcomes ($O_1,\dots,O_9$)”. The cohorts are:

- Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
  with at least one year of follow-up before treatment initiation and a recorded
  hypertension diagnosis within that year.
- Comparator cohort: Patients receiving any drug within the beta blocker class
  with at least one year of follow-up before treatment initiation and a recorded
  hypertension diagnosis within that year.
- Outcome cohorts: We considered 3 efficacy and 6 safety outcome cohorts. These
  were patients in the database with a diagnosis of: acute MI; hospitalization
  with heart failure; ischemic or hemorrhagic stroke (efficacy outcomes);
  hypokalemia; hyperkalemia; hypotension; angioedema; cough; abnormal weight
  gain (safety outcomes).

All cohort definitions were identical to the ones used in the multinational
study that provided overall treatment effect estimates comparing all
anti-hypertensive drug classes with each other [@Suchard2019]. More information
can be found in the supplementary material.

For our analyses we used data from 3 US claims databases, namely IBM MarketScan
Commercial Claims and Encounters (CCAE), IBM MarketScan Medicaid (MDCD), and IBM
MarketScan Medicare Supplemental Beneficiaries (MDCR). Our analyses included a
total of 784,561, 66,820 and 101,661 patients initiating treatment with ACE
inhibitors and 395,740, 45,999 and 69,798 patients initiating treatment with
beta blockers in CCAE, MDCD and MDCR respectively (Table 1). Adequate numbers of
patients were included in all strata of predicted acute MI risk.

@@

We internally developed separate prediction models for acute MI in all 3
databases. The prediction models were estimated on the propensity score matched
(1:1) subset of the sample, using caliper of 0.2 and after excluding patients
having the outcome any time prior to treatment initiation. We chose a time
horizon of 2 years after inclusion into the target cohort. We developed the
prediction models using LASSO logistic regression with 3-fold cross validation
for hyper-parameter selection.

The models developed in the 3 databases had moderate discriminative performance
(internally validated) with no major issues of overfitting to any cohort except
for the case of CCAE database in which the derived prediction model performed
better in the comparator cohort (Table 2). We also observed lower performance of
the prediction model developed in MDCR compared to the other 2 databases (see
supplementary material).

@@

We used patient-level predictions to stratify the sample into 4 MI risk
quarters. Within risk quarters, relative effects were estimated using Cox
regression and absolute effects were estimated from the Kaplan-Meier estimate
differences at 2 years after treatment initiation. To adjust for observed
confounding within risk strata, we estimated propensity scores using the same
approach as in the development of prediction models. We used the estimated
propensity scores to stratify patients into 5 strata, within each risk
quarter. The risk quarter-specific effect estimates were derived by averaging
over the estimates within the propensity score fifths.

There was sufficient overlap of propensity score distribution in all risk strata
(Figure 2). Propensity score adjustment achieved balance for most of the
considered covariates, measured using standardized mean differences before and
after adjustment (Figure 3). However, in lower risk strata imbalances persisted
for a substantial subset of the covariates, all related to pregnancy
findings. This was anticipated (use of ACE inhibitors is specifically
contraindicated during pregnancy) and was already pointed out in
[@Suchard2019]. Therefore, treatment effect estimates in the lowest risk
quarters of CCAE and MDCR may still be subject to residual confounding and
should be interpreted with care.

@@

```{r, echo=FALSE}
map_outcomes <- readRDS(here("data/processed/multipleRseeAnalyses/map_outcomes.rds"))
map_exposures <- readRDS(here("data/processed/multipleRseeAnalyses/map_exposures.rds"))
relative <- readRDS(here("data/processed/multipleRseeAnalyses/mappedOverallRelativeResults.rds")) %>%
  left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
  select(-estOutcome) %>%
  rename("estOutcome" = "outcome_name") %>%
  left_join(map_outcomes, by = c("stratOutcome" = "outcome_id")) %>%
  select(-stratOutcome) %>%
  rename("stratOutcome" = "outcome_name") %>%
  left_join(map_exposures, by = c("treatment" = "exposure_id")) %>%
  select(-treatment) %>%
  rename("treatment" = "exposure_name") %>%
  left_join(map_exposures, by = c("comparator" = "exposure_id")) %>%
  select(-comparator) %>%
  rename("comparator" = "exposure_name")

absolute <- readRDS(here("data/processed/multipleRseeAnalyses/mappedOverallAbsoluteResults.rds")) %>%
  left_join(map_outcomes, by = c("estOutcome" = "outcome_id")) %>%
  select(-estOutcome) %>%
  rename("estOutcome" = "outcome_name") %>%
  left_join(map_outcomes, by = c("stratOutcome" = "outcome_id")) %>%
  select(-stratOutcome) %>%
  rename("stratOutcome" = "outcome_name") %>%
  left_join(map_exposures, by = c("treatment" = "exposure_id")) %>%
  select(-treatment) %>%
  rename("treatment" = "exposure_name") %>%
  left_join(map_exposures, by = c("comparator" = "exposure_id")) %>%
  select(-comparator) %>%
  rename("comparator" = "exposure_name") %>%
  mutate(
    estimate = 100 * estimate,
    lower    = 100 * lower,
    upper    = 100 * upper
  )

acute_mi_relative <- relative %>%
  filter(
    stratOutcome == "acute_myocardial_infarction",
    estOutcome == "acute_myocardial_infarction"
  )

acute_mi_absolute <- absolute %>%
  filter(
    stratOutcome == "acute_myocardial_infarction",
    estOutcome == "acute_myocardial_infarction"
  )
acute_cough_relative <- relative %>%
  filter(
    stratOutcome == "acute_myocardial_infarction",
    estOutcome == "cough"
  )

acute_cough_absolute <- absolute %>%
  filter(
    stratOutcome == "acute_myocardial_infarction",
    estOutcome == "cough"
  )
```

For hospitalization with acute MI there was an increasing trend in favor ACE
inhibitors compared to beta blockers on the relative scale (hazard ratios
decreased) with increasing acute MI risk. More specifically, hazard ratios
decreased from 
`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(estimate)`
(`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(lower)` to
`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(upper)`;
95\% CI) and
`r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(estimate)`
(`r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(lower)` to
`r acute_mi_relative %>% filter(database == "mdcd", riskStratum == "Q1") %>% pull(upper)`;
95\% CI) to
`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(estimate)`
(`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(lower)` to
`r acute_mi_relative %>% filter(database == "ccae", riskStratum == "Q4") %>% pull(upper)`;
95\% CI),
